---
title: "Changepoint Analysis Using R"
author: "Robert Maidstone"
date: "17 October, 2018"
output:
  beamer_presentation:
      includes:
        in_header: ~/changepointsinR/header-beamer.tex
---



## What are Changepoints?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE,message=FALSE)
library(tidyverse)
library(patchwork)
library(qpcR)
library(changepoint)
library(prophet)
```


 \includegraphics[width = 10cm,clip]{def.jpg} 


## Change in mean

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(14)
m<-5
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
means <- rnorm(m+1,0,4)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  y[j]<-rnorm(length(j),means[i],1)
}

```
  \endcol
\begincol{.4\textwidth}
```{r pressure, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) + 
 # geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols


## Change in mean

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(14)
m<-5
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
means <- rnorm(m+1,0,4)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  y[j]<-rnorm(length(j),means[i],1)
}

```

```{r}
y_mean<-y
tru_cps_mean <- true_cps
```

  \endcol
\begincol{.4\textwidth}
```{r, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) + 
  geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols

## Change in variance

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(12)
m<-5
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
sd <- runif(m+1,1,20)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  y[j]<-rnorm(length(j),0,sd[i])
}

```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) +
  #geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols

```{r}
y_var<-y
tru_cps_var <- true_cps
```

## Change in variance

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(12)
m<-5
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
sd <- runif(m+1,1,20)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  y[j]<-rnorm(length(j),0,sd[i])
}

```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols

## Change in trend

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(110)
m<-3
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
slope <- rnorm(m+1,0,.01)
intercept <- rnorm(1,0,1)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  if(i==1){
  for(jind in j){
    y[jind]<-intercept+(jind-true_cps[i])*
      slope[i] + rnorm(1,0,1)
  }
  }else{
    for(jind in j){
      y[jind]<-y[j[1]-1]+(jind-true_cps[i])*
        slope[i] + rnorm(1,0,1)
    }
  }
}

```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) +
  #geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols

```{r}
y_trend<-y
tru_cps_trend <- true_cps
```

## Change in trend

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
set.seed(110)
m<-3
n<-1000
true_cps <- c(0,sort(sample(1:(n-1),m)),n)
slope <- rnorm(m+1,0,.01)
intercept <- rnorm(1,0,1)
y<-c()
for(i in 1:(m+1)){
  j <- (true_cps[i]+1):true_cps[i+1]
  if(i==1){
  for(jind in j){
    y[jind]<-intercept+(jind-true_cps[i])*
      slope[i] + rnorm(1,0,1)
  }
  }else{
    for(jind in j){
      y[jind]<-y[j[1]-1]+(jind-true_cps[i])*
        slope[i] + rnorm(1,0,1)
    }
  }
}

```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=4,fig.width=3}
ggplot(data.frame(x=1:n,y=y),aes(x=x,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=true_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()
```
  \endcol
\endcols

## Real World Examples

\begincols
  \begincol{.4\textwidth}
    \includegraphics[width = 5cm,clip]{copynumber.png} \
    \scriptsize (a) Copy number at genomic positions in a human breast tumor sample (Chen & Wang, 2009). 
  \endcol
  \begincol{.4\textwidth}
       \includegraphics[width = 5cm,clip]{BFM.png} \
  \scriptsize (b) Rotation of a bacterial flagella motor (Maidstone, 2016).
  \endcol
\endcols

## Many changepoint methods

Many different changepoint algorithms exist
\begincols
  \begincol{.4\textwidth}
  
  - Exhaustive Search
  - Optimal Partitioning
  - PELT
  - FPOP (and R-FPOP)
  - CROPS
  - Segment Neighbourhood Search
  - pDPA
  - SNIP
  
\endcol
\begincol{.4\textwidth}

  - Binary Segmentation
  - WBS
  - CBS
  - SMUCE
  - SMOP
  - ED-PELT
  - E-Divisive
  - ECP
  
\endcol
\endcols


## Binary Segmentation

\begin{block}{}
\small{
\begin{description}
\item[Input:] A set of data of the form, $(y_1,y_2,\hdots,y_n)$ where $y_i\in\RR$.\\
A test statistic on the data $\Gamma(\cdot)$, \\
An estimator of changepoint location $\hat{\tau}(\cdot)$,\\
A rejection threshold $c$.
\item[Initialise:] Let $\mathcal{CP}=\emptyset$ and $\mathcal{S}=\{[1,n]\}$;
\item[Iterate:] while $\mathcal{S}\neq\emptyset$ do
\begin{itemize}
\item Choose an element of $\mathcal{S}$; denote this element as $[s,t]$;
\item if $\Gamma(\mathbf{y}_{s:t})<c$ then
\begin{itemize}
\item remove $[s,t]$ from $\mathcal{S}$
\end{itemize}
\item if $\Gamma(\mathbf{y}_{s:t})\geq c$ then
\begin{itemize}
\item remove $[s,t]$ from $\mathcal{S}$;
\item calculate $r=\hat{\tau}(\mathbf{y}_{s:t})+s-1$, and add $r$ to $\mathcal{CP}$;
\item if $r \neq s$ add $[s,r]$ to $\mathcal{S}$;
\item if $r \neq t-1$ add $[r+1,t]$ to $\mathcal{S}$;
\end{itemize}
\end{itemize}
\item[Output:] The changepoints recorded in $\mathcal{CP}$.
\end{description}}
\end{block}

## Binary Segmentation using "wbs"

```{r}
y <- y_mean
tru_cps <- tru_cps_mean[-c(1,length(tru_cps_mean))]
```

\begincols
  \begincol{.6\textwidth}
\footnotesize
Baranowski and Fryzlewicz (2015)
\scriptsize
\vspace{20pt}

```{r,echo=TRUE}
library(wbs)

sbs(y) -> sbs.out
s.cpt <- changepoints(sbs.out,th =4)

s.cpt$cpt.th[[1]] %>% sort

wbs(y,
    M=5000) -> wbs.out

w.cpt <- changepoints(wbs.out,th=4)

w.cpt$cpt.th[[1]] %>% sort

```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y_mean),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=tru_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()+ggtitle("True")
```

```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y_mean),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=s.cpt$cpt.th[[1]]),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle("BS")
```

```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y_mean),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=w.cpt$cpt.th[[1]]),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle("WBS")
```
  \endcol
\endcols


## Cost function representation

Most changepoint detection methods boil down to minimising the sum of some cost function, $\mathcal{C}(\cdot)$, over the segments.

\[
\min_{\tau_{1:m},m} \left[\sum_{j=0}^{m+1} \mathcal{C}(\mathbf{y}_{\tau_j+1:\tau_{j+1}})\right]
\]
This cost function could be a number of things:

1. Negative log-likelihood,
2. Negative posterior,
3. Minimum Description Length.

## Dynamic Programming Methods

\begin{description}
\item[Optimal Partitioning:] Optimisation based sum of optimal up to last changepoint and the cost between last changepoint and current time (plus a penalty to avoid over fitting).
\begin{block}{}
\setlength\abovedisplayskip{0pt}
\begin{align*}
F(\tau^*)=\min_{0\leq\tau<\tau^*}[F(\tau)+\mathcal{C}(\mathbf{y}_{(\tau+1):\tau^*})+\beta].
\end{align*}
\end{block}
\item[Segment Neighbourhood Search:] Optimisation for $k$ segments based on optimal for $k-1$ segments plus cost for new segment.
\begin{block}{}
\setlength\abovedisplayskip{0pt}
\begin{align*}
q_{1,j}^k=\min_{v\in \{1,\hdots,j-1\}}[q_{1,v}^{k-1}+q_{v+1,j}^1].
\end{align*}
\end{block}
\end{description}

## The changepoint package

- Functions for changepoint analysis.
- Can use various changepoint detection methods:
     * Binary Segmentation
     * AMOC
     * Segment Neighbourhood
     * PELT (Optimal Partitioning)
- With various choices of penalty function/value.
- And either a Gaussian or CUSUM test statistic.
- Three *headliner* functions:
     * `cpt.mean`
     * `cpt.var`
     * `cpt.meanvar`
- Authors: Rebecca Killick, Kaylea Haynes, Idris Eckley, Paul Fearnhead, Jamie Lee.

## Change in mean using "changepoint"

```{r}
tru_cps <- tru_cps_mean[-c(1,length(tru_cps_mean))]
```

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
changepoint::cpt.mean(y_mean,
                      penalty = "BIC",
                      method = "PELT"
                      ) -> cpt_object #S4

cpt_object %>% changepoint::likelihood()
cpt_object %>% changepoint::ncpts()
cpt_object %>% changepoint::cpts() -> est_cps
qpcR:::rbind.na(tru_cps,est_cps)
```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=3,fig.width=3}
ggplot(data.frame(t=1:n,y=y_mean),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=tru_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()+ggtitle("True")
```

```{r, fig.height=3,fig.width=3}
ggplot(data.frame(t=1:n,y=y_mean),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=est_cps),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle("Estimated")
```

  \endcol
\endcols

## Change in variance using "changepoint"

```{r}
tru_cps <- tru_cps_var[-c(1,length(tru_cps_var))]
```


\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r,echo=TRUE}
changepoint::cpt.var(y_var,
                     penalty = "BIC",
                     method = "PELT"
                     ) -> cpt_object #S4

cpt_object %>% changepoint::likelihood()
cpt_object %>% changepoint::ncpts()
cpt_object %>% changepoint::cpts() -> est_cps
qpcR:::rbind.na(tru_cps,est_cps)
```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=3,fig.width=3}
ggplot(data.frame(t=1:n,y=y_var),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=tru_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()+ggtitle("True")
```

```{r, fig.height=3,fig.width=3}
ggplot(data.frame(t=1:n,y=y_var),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=est_cps),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle("Estimated")
```

  \endcol
\endcols

## Penalty Choice

\begin{description}
\item[Optimal Partitioning:] Optimisation based sum of optimal up to last changepoint and the cost between last changepoint and current time (plus a penalty to avoid over fitting).
\begin{block}{}
\setlength\abovedisplayskip{0pt}
\begin{align*}
F(\tau^*)=\min_{0\leq\tau<\tau^*}[F(\tau)+\mathcal{C}(\mathbf{y}_{(\tau+1):\tau^*})+\beta].
\end{align*}
\end{block}
\end{description}

## Penalty Choice

\begincols
  \begincol{.5\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r}
y<-y_mean
tru_cps <- tru_cps_mean[-c(1,length(tru_cps_mean))]
```


```{r, fig.height=2,fig.width=3,results="hide"}
cpt.mean(y,penalty = "CROPS",
         pen.value=c(5,1000),
         method="PELT",
         class=FALSE) -> crops.out

ggplot(data.frame(t=1:n,y=y),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=tru_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()+ggtitle("True")
```

```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=crops.out$changepoints[[6]]),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle(paste0("beta=",round(crops.out$cpt.out[1,6],2)))
```
  \endcol
\begincol{.5\textwidth}

```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=crops.out$changepoints[[9]]),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle(paste0("beta=",round(crops.out$cpt.out[1,9],2)))
```

```{r, fig.height=2,fig.width=3}
ggplot(data.frame(t=1:n,y=y),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=crops.out$changepoints[[1]]),aes(xintercept=CP),colour="blue",linetype="dashed") +
  theme_bw()+ggtitle(paste0("beta=",crops.out$cpt.out[1,1]))
```

  \endcol
\endcols

## CROPS




\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize
```{r}
y<-y_mean
```


```{r, echo=TRUE,message=FALSE}
cpt.mean(y,penalty = "CROPS",
         pen.value=c(5,1000),
         method="PELT",
         class=FALSE) -> crops.out
```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=3,fig.width=3}
crops.out$cpt.out %>% t() %>% as.data.frame() %>% ggplot(aes(numberofchangepoints,penalised_cost)) + geom_line() +geom_point() +theme_bw()
```

  \endcol
\endcols

\scriptsize
```{r, echo=TRUE}
crops.out$cpt.out %>% .[,1:5]
```



## Change in trend with Prophet


\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize

```{r}
y<-y_trend
tru_cps <- tru_cps_trend
```

```{r,echo=TRUE, eval=FALSE}
df <-data.frame(
  ds=lubridate::as_date("16/10/18")+
                        (1:length(y)),
  y= y)

m <- prophet(df)

forecast <- predict(m)

plot(m, forecast)+add_changepoints_to_plot(m)

prophet_plot_components(m, forecast)
```
  \endcol
\begincol{.4\textwidth}
```{r, fig.height=3,fig.width=3}
ggplot(data.frame(t=1:n,y=y_trend),aes(x=t,y=y)) + geom_point(size=.5) +
  geom_vline(data=data.frame(CP=tru_cps),aes(xintercept=CP),colour="red",linetype="dashed") +
  theme_bw()+ggtitle("True")
```

```{r,echo=FALSE, eval=TRUE,fig.height=3,fig.width=3,message=FALSE,warning=FALSE,results="hide"}
df <-data.frame(ds=lubridate::as_date("16/10/18")+(1:length(y_trend)),y= y_trend)

m <- prophet(df)

forecast <- predict(m)

plot(m, forecast)+add_changepoints_to_plot(m)
```

  \endcol
\endcols

## Peyton Manning (Prophet example)

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize


```{r,echo=TRUE, eval=FALSE}
df<-read.csv(
  "~/changepointsinR/
  example_wp_log_peyton_manning.csv")
m <- prophet(df)

future <- make_future_dataframe(
  m,
  periods = 365)


forecast <- predict(m, future)

plot(m, forecast)+add_changepoints_to_plot(m)
```
  \endcol
\begincol{.4\textwidth}

```{r,echo=FALSE, eval=TRUE,fig.width=5,message=FALSE,warning=FALSE,results="hide"}
df<-read.csv(
"~/changepointsinR/example_wp_log_peyton_manning.csv"
)
m <- prophet(df)

future <- make_future_dataframe(
  m,
  periods = 365)


forecast <- predict(m, future)

plot(m, forecast)+add_changepoints_to_plot(m)
```

  \endcol
\endcols

## Peyton Manning (Prophet example)

```{r}
plot(m, forecast)+add_changepoints_to_plot(m)
```


## Peyton Manning (Prophet example)

\begincols
  \begincol{.6\textwidth}
<!-- \footnotesize -->
\scriptsize


```{r,echo=TRUE, eval=FALSE}
prophet_plot_components(m, forecast)
```
  \endcol
\begincol{.4\textwidth}

```{r,echo=FALSE, eval=TRUE,fig.width=4,message=FALSE,warning=FALSE,results="hide"}
prophet_plot_components(m, forecast)
```

  \endcol
\endcols

## Discussion

- Many changepoint methods avaliable
      * Not discussed; Multivariate, online changepoint detection, non parametric (`changepoint.np`,`ecp`) and more.
      
- `changepoint` and `wbs` packages are a good place to start.
- Computational time can be a major issue in changepoint detection, meaning bespoke solutions can be best.
- Code not always in the best form, but big forcasting packages such as prophet use changepoints too.
      

